Surface Source Extension, Reflection Case

### 2019-12-12 repro/papers/sserefl/paper.tex
many edits

In the following sections, I will first give a more precise statement
of the main result. Then I will review plane-wave modeling, and construct
an asymptotic approximation to single-reflector plane wave data. In
the final section I will present the surface source extended FWI
objective and compute its gradient at a homogeneous background with
single reflector data. The main result will follow immediately from
this calculation.

@@

\section{Main Result}

We adopt the usual pressure-velocity acoustics system as the
underlying physics of waves, see beginning of next section. The
surface source extension allows the source to extend over the surface,
in the form $w(x,t)\delta(z)$ for a nominal source position of
$(0,0)$. After Radon transform, this source takes the form
$\tw(\xi,t)\delta(z)$, where $\xi$ is the plane wave slowness.

@@

\section{Plane Wave Modeling}
Start with 2D acoustics in velocity-pressure form: $p$ is excess
pressure, $\bv=(v_x,v_z)$ is particle velocity about equilibrium, $f$
is a (possibly) extended source model, all linked by the acoustic wave system
\begin{eqnarray}
\label{eqn:awe}
\frac{\partial p}{\partial t} + \kappa \nabla \cdot {\bf v} &=&
                                                                f(\bx,t;\bx_s) \nonumber\\
\rho \frac{\partial {\bf v}}{\partial t} + \nabla p = 0 \nonumber\\
p,{\bf v}&=&0, t \ll 0
\end{eqnarray}

As in other work on this topic, I will assume $\rho$ to be fixed, and
for sake of simplicity constant, and focus the discussion on
$\kappa$. Thus the modeled pressure traces at receiver locations
depend on $\kappa$ and the source function $f$. For simplicity again, assume that
receivers are located at every point of the hyperplane $z=z_r$. Then
the modeling operator is defined as
\begin{equation}
\label{eqn:fwd}
S[\kappa]f = p|_{z=z_r}
\end{equation}

Assume that sources all lie on the surface $z=z_s$, so that surface
extended sources take the form $f(\bx,t;\bx_s) =
w(x,t)\delta(z-z_s)$. By abuse of notation, regard $S[\kappa]$ as
acting on $w$.

@@

\append{1D Wave Motion}
Begin with the 1D acoustics point source system. 
\begin{eqnarray}
\label{eqn:awe1d}
\frac{\partial p}{\partial t} +\kappa\frac{\partial 
  v}{\partial z} &=& w(t)\delta(z-z_s) \nonumber\\
\rho \frac{\partial v}{\partial t} + \frac{\partial p}{\partial 
  z}&=&0\nonumber\\
 p,v&=&0, t \ll 0. 
\end{eqnarray}
Since the right hand side is singular, so is the solution, so it must
be a solution in the weak sense; since it is causal, it is uniquely
determined by the system above. It follows from the weak solution
conditions that the pressure is continuous at $p=0$, whence $v$ must
have a discontinuity.

Now assume that $\kappa, \rho$ are constant. 
In $z \ne 0$, the right hand side 
vanishes, so the solution must be locally a combination of plane
waves; causality implies that
\[
p(z,t)=a\left(t \mp \frac{z}{c}\right), \, v(z,t)=\pm b\left(t \mp
  \frac{z}{c}\right)
\]
From the second dynamical equation (Newton's law) it follows that $b =
a/(\rho c)$. The singularity on the LHS of the first dynamical
equation (constitutive law) is
\[
\kappa [v]_{z=0}\delta(z) =
2\frac{\kappa}{c}b\delta(z) = 2a\delta(z).
\] 
This must in turn equal the RHS of the constitutive law, whence
$a=w/2$. Thus
\begin{eqnarray}
\label{eqn:sol1d}
p(z,t) &=& \frac{1}{2}w\left(t \mp \frac{z}{c}\right) \nonumber \\
v(z,t) &=& \pm\frac{1}{2\rho c}w\left(t \mp \frac{z}{c}\right)
           \nonumber \\
\pm z &>& 0.
\end{eqnarray}

@@

If $\kappa, \rho $ are constant, it is possible to develop explicit
expressions for the solution of this system. Assuming that $\delta
\rho=0$ and $\delta \kappa$ is nonzero only in $z>0$, the reflected
(perturbational) wavefield should be upcoming (i.e. a function of
$z+ct$, away from the wavefront), as the wave scatters
only once, at the wavefront implicit in the expressions
\ref{eqn:sol1d}. On the other hand, the field must vanish before the
first arrival.

>>>>

The first of the following sections sets out the framework of this
study. It does {\em not} include mathematically complete definitions
or fully qualified statements of results - these may be found in their
entirety in later sections. Instead, it is intended to introduce the
concepts that figure in the main results, which are stated succinctly
in the third section. The  I will review plane-wave modeling, and construct
an asymptotic approximation to single-reflector plane wave data. In
the final section I will present the surface source extended FWI
objective and compute its gradient at a homogeneous background with
single reflector data. The main result will follow immediately from
this calculation.

@@

\section{Preliminaries}

This paper will use linear acoustics in velocity-pressure form as its
model for wave motion. In this model, the space-time fields $p$ is excess
pressure, $\bv$ is particle velocity about equilibrium, $f$
represents an energy source as a defect in the acoustic constitutive
law. The bulk modulus $\kappa$ and density $\rho$ are positive functions of
spatial position. The acoustic wave system couples these quantities:
\begin{eqnarray}
\label{eqn:awe}
\frac{\partial p}{\partial t} + \kappa \nabla \cdot {\bf v} &=&
                                                                f \nonumber\\
\rho \frac{\partial {\bf v}}{\partial t} + \nabla p = 0 \nonumber\\
p,{\bf v}&=&0, t \ll 0
\end{eqnarray}
In terms of the solution of the system \ref{eqn:awe}, the modeling
operator is defined by
\begin{equation}
\label{eqn:fwd}
S[\kappa]f = p|_{z=z_r}
\end{equation}
Two special types of right-hand side are important in the discussion
to follow: isotropic point sources, of the form
\begin{equation}
  \label{eqn:ipr}
  f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s).  
\end{equation}
and surface sources, of the form
\begin{equation}
  \label{eqn:ss}
  f(\bx,t;\bx_s) = u(x,y,t;x_s,y_s)\delta(z-z_s). 
\end{equation}
Point sources are a special case of surface sources. Introduce appropriately
parametrized versions of the modeling operator for these two cases:
\begin{eqnarray}
 f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s) & \Rightarrow & S[\kappa]f = S_p
  [\kappa]w\\ \nonumber
  \label{eqn:specsrc}
f(\bx,t;\bx_s) = u(x,y,t;x_s,y_s)\delta(z-z_s) & \Rightarrow & S[\kappa]f =
  S_s[\kappa]u
\end{eqnarray}

The standing assumption in this work is that observed data to be fit
is to good approximation the result of a point source. That is, there
are bulk modulus $\kappa_*(\bx)$ and source wavelet or pulse $w_*(t)$
so that
\begin{equation}
  \label{eqn:datadef}
  d = S_p[\kappa_*]w_*
\end{equation}
In equation \ref{eqn:datadef}, $d$ is a function of receiver position
$\bx_r$, source position $\bx_s$, and time $t$.

Full waveform inversion (FWI), as construed in this paper, means:
\begin{equation}
  \label{eqn:deffwi}
  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }w \mbox{ so that }d
  \approx S_p[\kappa]w.
\end{equation}
In the following paragraphs, I will discuss four approaches to FWI:

\subsection{``Straight'' FWI}

The standard approach to FWI is to formulate the task \ref{eqn:deffwi}
as a nonlinear least squares problem:
\begin{equation}
  \label{eqn:defols}
  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }w \mbox{ to
    minimize }
  \| d -  S_p[\kappa]w\|^2.
\end{equation}
Here the symbol $\|\cdot\|^2$ signifies {\em mean-square}, that is,
sum the squares of every value of the field between the vertical bars.

This concept was introduced in the 1980's
(\cite{BamChavLai:79,Tara:84a,KolbColLai:86,Crasetal:90},
and many other papers since then). Usually some form of regularization
is applied, to compensate for poorly determined aspects of $\kappa$
and/or $w$, as is explained in Tarantola's influential book
\cite[]{Tarantola:05}. The approach was of only academic interest
until roughly fifteen years ago, when 3D modeling and related computations
first became feasible at industrial scale, and evidence began to emerge 
that FWI offered potentially decisive improvements in imaging accuracy
over other inversion (imaging) methods, especially for construction of
large-scale earth model features
\cite[]{Plessix:10,Vigh:10}. In the last ten years, FWI has become
both a production service offered by the geophysical industry, and a
leading topic of both industry-oriented and academic research.

As was understood very early in the evolution of FWI, the basic
least-squares formulation \ref{eqn:defols} suffers from a major
defect: the parameter spaces tend to be so large that only iterative
optimization (that is, a variant
of Newton's method) may feasibly be applied, due to computational cost
- this remains true today and for the foreseeable future - but these
method tend not to produce usable results unless the initial iterate
is in some sense ``good'' (predicts travel times to within a
half-wavelength) \cite[]{GauTarVir:86,Plessix:10}.``Good'' initial models are not always
available. Accordingly, a number of alternative approaches have been
proposed, many of them variations on the least squares theme, amongst
which are the three approaches described below.

\subsection{Source Extensions}

Careful analysis of the defect mentioned in the last paragraph shows
that the failure of iterative optimization applied to \ref{eqn:defols}
to converge (or at least to converge usefully quickly) appears to
arise from failure to fit the data: some parts of the oscillatory time
series $d(\cdot,t)$ overlap oscillations in $S_p[\kappa]w$, while
others do not. It seems reasonable to guess that this ``cycle skipping'' phenomenon 
could be avoided if the data could be fit throughout the inversion
process. That requirement in turn implies a modeling operator with
more degrees of freedom, that is, a bigger domain, than the basic
modeling operator $S_p$, which should produce the same results as
$S_p$ when applied to pairs of bulk moduli and isotropic point sources \ref{eqn:ipr}.

There is
well-established mathematical terminology for this relationship:
operator A is an {\em extension} of operator B if (i) the domain of $A$
contains the domain of $B$, and whenever $v$ is in the domain of $B$,
$Av = Bv$. Two extensions of $S_p$ already appear
above: the isotropic point radiator \ref{eqn:ipr} is a special case of
both the surface source \ref{eqn:ss}, which is in turn a special case
of the generic right-hand side in the system \ref{eqn:awe}. So the
domain of $S_s$ contains the domain of $S_p$, as does the domain of
$S$. From the relations \ref{eqn:specsrc}, $S$ and $S_s$ produce the
same result as $S_p$ when applied to sources in their common domain,
so these operators are both extensions of $S_p$.

Since these extensions of $S_p$ involve additional degrees of freedom
for the energy source in the system \ref{eqn:awe}, they have come to
be called {\em source extensions}. These are not the only source
extensions - see \cite{HuangNammourSymesDollizal:SEG19} for an
overview of source extensions and their properties. Extensions may
also be constructed by adding degrees of freedom to the coefficients
in the wave system (bulk modulus, in the present case), 
rather than the right hand side - see \cite{geoprosp:2008} for a
description of some of these {\em medium extensions}.

The purpose of any extension is to add options for data fitting,
thereby reducing the chances of cycle-skipping. As will be seen below,
both extension $S$ and $S_s$ permit arbitrarily accurate fitting of
essentially any data with essentially any bulk modulus field $\kappa$,
with proper choice of $f$ or $w$ respectively. However, the standing
assumption on the data $d$ is that it derives from a point source, per
\ref{eqn:datadef}. So the added degrees of freedom represented by the
extensions $S$ and $S_s$ must be suppressed somehow. As explained by
\cite{geoprosp:2008}, a convenient mechanism for suppression of the
extension degrees of freedom is a penalty term added to the objective
of the extended
version of the least squares problem \ref{eqn:defols}. Choice of penalty
term, along with choice of extension, characterizes each of the extended
inversion methods detailed in the next few paragraphs.

\subsubsection{Wavefield Reconstruction Inversion}
This concept was introduced by \cite{LeeuwenHerrmannWRI:13}, and further
developed by \cite{LeeuwenHerrmann:16,WangYingst:SEG16} and other
authors. It is based on the extension $S$ (equation \ref{eqn:fwd}), with full
space-time volume source $f$, and presumes that the correct point
source wavelet $w_*$ is known. 
\begin{equation}
  \label{eqn:defwri}
  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }f \mbox{ to
    minimize }
  \| d -  S[\kappa]f\|^2+\lambda^2\|f-w_*\delta(\cdot-\bx_s)\|^2.
\end{equation}
Evidently the difference $f-w_*\delta(\cdot-\bx_s)$ should be such
that the scaled sum-of-squares makes sense - this is always the case
for discretizations, but for the continuum limit entails an actual
assumption.

\cite{LeeuwenHerrmannWRI:13} posed this problem slightly
differently, in terms of the second order wave equation for the
pressure wavefield $p$. From this viewpoint, the 
second terms becomes a scaled mean square of the residual in the
wave equation. The formulation presented here is equivalent, and was
introduced by \cite{WangYingst:SEG16}. \cite{LeeuwenHerrmann:16} used
the variable projection method \cite[]{GolubPereyra:03}, eliminating the source $f$ in the
inner step and updating the bulk modulus (or an equivalent quantity)
in the outer step. In any penalty method, control of the penalty
parameter has a large influence on the speed of
convergence. \cite{Aghamiry:19} use an augmented Lagrangian algorithm
to minimize the influence of the penalty weight choice.

The Achilles' heel of Wavefield Reconstruction Inversion (WRI), in any of
the formulations described so far, is the necessity of storing a full
space time field, namely $f$ (or $p$). Especially for time-domain
formulation and high frequency waves, this storage requirement can be
formidable. In fact, ``most'' source fields $f$ are non-radiating,
that is, $S[\kappa]f=0$. The orthocomplement of this subspace of
non-radiating sources is the range of the adjoint operator
$S[\kappa]^T$. Therefore the sources may be re-parametrized as images of
$S[\kappa]^T$. Write $f = g + w_*\delta(\cdot-\bx_s)$: then the second term in
the objective displayed in \ref{eqn:defwri} is simply $\|g\|^2$.
Decompose $g = S[\kappa]^Te + n$, in
which $e$ is the same type of object as $d$ and $S[\kappa]n=0$ (that
is, $n$ is a non-radiating
source), and note that the decomposition is orthogonal. Then the WRI problem \ref{eqn:defwri} may be reformulated as
\begin{equation}
  \label{eqn:defwrireform}
  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }e \mbox{ to
    minimize }
  \| d -  S[\kappa]w_*\delta(\cdot-\bx_s) - S[\kappa]S[\kappa]^Te\|^2+\lambda^2\|S[\kappa]^Te\|^2.
\end{equation}
Note that $n$ has disappeared: {\em a priori} it does not contribute to the first
term, and $n \ne 0$ makes the second term larger. So minimization
necessarily entails the choice $n=0$.

This reformulation does {\em not} require storage of $d+1$-dimensional
fields, as does the original WRI formulation \ref{eqn:defwri}. The
field $e$ has taken the place of the space-time source $f$, and its
size is precisely the same as that of the data field $d$. Therefore
the extra memory cost over modeling is the same as another copy of
data. A close examination reveals that in computing the composition
$SS^T$ via time-stepping methods, the same simultaneity issue arises
as in all applications of the adjoint state method. However good
solutions are known \cite[]{Griewank:book}, and since precisely the
same issue arises in time-domain FWI, the cost is the same. Therefore,
it is this reformulation of WRI that will be analyzed below.

\subsubsection{Space-Time Extension}
\cite{HuangSymes:Geo18a} introduced a variant of WRI, also using the space-time extension, but not
requiring knowledge of the target point source wavelet $w_*$: in fact,
an estimate of this wavelet is a by-product of the algorithm. This
approach relies on an {\em annihilator}, a linear operator whose null
space is precisely the desired class of source models, in this case
isotropic point radiators. Such an operator can be used to generate a
quadratic form whose zero set is also the desired class, hence can be
used as a penalty in a nonlinear least squares objective, designed to
drive the source towards to the desired type. See \cite{geoprosp:2008}
for a discussion of annihilators in extended inversion.

The annihilator proposed by \cite{HuangSymes:Geo18a} is multiplication
by $|\bx-\bx_s|$, that is,
\begin{equation}
  \label{eqn:defann}
  Af(\bx,t;\bx_s) = |\bx-\bx_s|f(\bx,t;\bx_s).
\end{equation}
The space-time extension approach consists of solving the optimization
problem:
\begin{equation}
  \label{eqn:defste}
  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }f \mbox{ to
    minimize }
  \| d -  S[\kappa]f\|^2+\lambda^2\|Af\|^2.
\end{equation}
This problem can be reformulated in the same way as the WRI problem. Write
$f = w\delta(\cdot-\bx_s) + S[\kappa]^Te + n$, where $n$ is
non-radiating as before: then the problem \ref{eqn:defste} is
equivalent to
\[
   \mbox{Given } d, \mbox{ find }\kappa, w, e \mbox{ and }n \mbox{ to
     minimize }
 \]
\begin{equation}
  \label{eqn:defstereform}
  \| d -  S[\kappa]w_*\delta(\cdot-\bx_s) - S[\kappa]S[\kappa]^Te\|^2+\lambda^2\|A(S[\kappa]^Te+n)\|^2.
\end{equation}
Note that the non-radiating component $n$ must remain in the second
term of the objective, because there is no reason for $n$ to be
$A^TA$-orthogonal to the range of $S[\kappa]^T$. Also, the isotropic
point radiator component $w\delta(\cdot-\bx_s)$ must also remain as
part of the unknowns - simply because it is unknown! It does not
appear in the second term because $A$ annihilates it.

\subsubsection{Surface Source Extension}
Beside the reformulation of WRI described above, it is also possible
to avoid the storage problem of WRI by limiting the support of the
extended source $f$ to lower dimensional sets. Of several
possibilities discussed in the literature, I will discuss the 
  surface source extension approach introduced by
\cite{HuangNammourSymesDollizal:SEG19}. The extension in question is
defined in equation \ref{eqn:defss}, and denoted $S_s$. The same
annihilator \ref{eqn:defann} is used in this approach, this time
interpreted as acting in the plane $z=z_s$, that is,
\[
  Au(x,y,t;x_s,y_s) = \sqrt{(x-x_s)^2+(y-y_s)^2}u(x,y,t;x_s,y_s).
\]
The problem formulation is
\begin{equation}
  \label{eqn:defsse}
  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }u \mbox{ to
    minimize }
  \| d -  S_s[\kappa]u\|^2+\lambda^2\|Au\|^2.
\end{equation}
Evidently the extended source has as many parameters as the data, so
the memory complexity of this problem is comparable to that of FWI or
reformulated WRI. The computational complexity compares favorably to
that of reformulated WRI.

\subsection{Layered Media, Plane Waves}
As explained in the introduction, this study explores the properties
of the various FWI variants explained above for the special case of
layered media: $\kappa = \kappa(z)$,
$\rho=\rho(z)$. Then the modeled shots differ from
each other by a horizontal translation, so that there is in effect
only one shot, for which I take the location to be
$\bx_s=(0,0,0)$ and suppress $\bx_s$ from the notation.

If $\log \kappa$ and $\log \rho$ are globally bounded and $f$ has
compact support (and some other properties that will be made precise
later), there is $\xi_{\rm max} > 0$ so that for
$\|\xi\|<\xi_{\rm max}$, the integrals defining slant stack fields
$\tilde{p},\tilde{\bv}$, namely
\begin{eqnarray}
\label{eqn:slant}
\tp(\xi,z,t)& = &\int dx\, dy\, p(x,y,z,t+\xi x),\nonumber\\
\tv(\xi,z,t)&=&\int dx\, dy\, v_z(x,y,z,t+\xi x),\nonumber\\
\tf(\xi,z,t)&=&\int dx\,dy\, f(x,y,z,t+\xi x),
\end{eqnarray}
are absolutely convergent, and $\tp, \tv$ satisfy the plane wave
version of the acoustic system \ref{eqn:awe}:
\begin{eqnarray}
\label{eqn:awepw}
(1-c^2\xi^2)\frac{\partial \tp}{\partial t} + \kappa\frac{\partial
  \tv}{\partial z} &=& \tf \nonumber\\
\rho \frac{\partial \tv}{\partial t} + \frac{\partial \tp}{\partial
  z}&=&0\nonumber\\
\tp,\tv&=&0, t \ll 0.
\end{eqnarray}
Here $c=\sqrt{\kappa/\rho}$ is the sound velocity. I use $\xi$ instead
of the more usual $p$ for plane wave slowness since $p$ already has
been assigned to mean pressure. Note that for a isotropic point
source, respectively a surface source,
\begin{eqnarray}
  f(\bx,t;\bx_s)=w(t)\delta(\bx) & \Rightarrow &
                                               \tf(\xi,z,t)=w(t)\delta(z)\\ \nonumber 
  f(\bx,t;\bx_s) = u(x,y,t)\delta(z) & \Rightarrow &
                                                     \tf(\xi,z,t) =
                                                     \tu(\xi,t)\delta(z)\\
  \nonumber
  \tu(\xi,t) &=& \int dx\,dy\, u(x,y,t+\xi x)
\end{eqnarray}
The plane wave domain modeling
operator for space-time extended sources samples $\tp$ on $z=z_r$:
\begin{equation}
\label{eqn:modpw}
\tS[\kappa]\tf = \tp|_{z=z_r}
\end{equation}
Similarly, the plane wave domain versions of $S_p$ and $S_s$ are
\begin{eqnarray}
 \tf(\xi,z,t) = w(t)\delta(z) & \Rightarrow & \tS[\kappa]\tf = \tS_p
  [\kappa]w\\ \nonumber
  \label{eqn:specsrc}
\tf(\xi,z,t) = \tu(\xi,t)\delta(z-z_s) & \Rightarrow & \tS[\kappa]\tf =
  \tS_s[\kappa]\tu
\end{eqnarray}

For homogeneous $\kappa=\kappa_0, \rho=\rho_0, c=c_0$, a computation
recapitulated in the Appendix C, relying on computations for the 1D
acoustics system detailed in Appendices A and B, yields
\begin{equation}
\label{eqn:modhom}
\tS_s[\kappa_0]\tu(\xi,t) = \frac{1}{2(1-c_0^2\xi^2)}\tu
\left(\xi,t-z_r\sqrt{\frac{1}{c_0^2}-\xi^2}\right).
\end{equation}
That is, at a homogeneous model, the plane wave modeling operator is
simply scaling by a $\xi$-dependent factor.

\section{Main Results}

The space dimension is not terribly important in the results to
follow. Generally I will assume it to take the physical value $d=3$,
though the numerical examples will use $d=2$ for reasons of
computational cost. For spatial position, write $\bx = (x,y,z)$ (or $\bx = (x,z)$ for
2D).

\cite{BlazekStolkSymes:13} give conditions under which the system
\ref{eqn:awe} has a strong solution, that is, in which all of the
terms in equations above are interpretable as functions of space-time. Assume that

\indent Assumption A: $\log \kappa, \log \rho$ are bounded and
measureable. \\

\indent Assumption B: $f$ is square-integrable, has a square-integrable time
  derivative, and \\has compact support.\\

Then Theorem 1 in \cite{BlazekStolkSymes:13} implies that there exists
a unique causal (vanishing for $t \ll 0$) strong solution
$(p,\bv)$ of the system \ref{eqn:awe}, that is, $p$ and $\bv$ are
$C^1$ functions of time with square-integrable (in space) values,
$\nabla p$ and $\nabla \cdot \bv$ are $C^0$ functions of time with
square-integrable values,  and the equations \ref{eqn:awe} are
satisfied pointwise a. e.

The central object in this study is the relation between the
coefficients $\kappa$ and $\rho$, on the one hand, and the fields $p$
and $\bv$ on the other. The influence of the wave velocity $c$ is the
most important, so from here on $\rho$ will be fixed and constant:
$\rho(\bx) = \rho_0 \in \bR^+$. Thus view $p$ and $\bv$ as functions of
$\kappa$ and $f$.

In seismic and ultrasonic experiments, the acoustic
fields are recorded at or near a hypersurface (the surface of the
earth, or the skin, or...). Theorem 2 in \cite[]{BlazekStolkSymes:13}
assures that the trace of $p$ on a cylindrical hypersurface $\Gamma
\times [0,T]$ is well-defined, as is the trace of the normal component
of $\bv$. These traces will serve as idealized data in the discussion
to follow, ignoring issues of sampling and other details of actual
measurement processes.

This discussion further idealizes acquisition by setting $\Gamma =
\{\bx: z=z_r\}$. This is not a bad approximation for some field
settings, and is convenient for the eventual use of the Radon
transform.

Active (artificial) sources (right-hand sides in the wave equations)
are frequently localized in space, relative to other active length scales. The
simplest such localized source is the {\em isotropic radiator},
\begin{equation}
  \label{eqn:ipr}
  f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s).  
\end{equation} 
in which $\bx_s$ is the location of the radiator. The radiator is a
special case of the {\em surface source}, in which the support of the
right-hand side in \ref{eqn:awe} is confined to a hypersurface. For
present purposes, it suffices to consider sources localized on the
hyperplane $z=z_s$:
\begin{equation}
  \label{eqn:ss}
  f(\bx,t;\bx_s) = w(x,y,t)\delta(z-z_s). 
\end{equation}
Neither distribution defined by \ref{eqn:ipr}, \ref{eqn:ss} is 
square-integrable, so the theory in \cite[]{BlazekStolkSymes:13} does
not apply directly to the system \ref{eqn:awe} with $f$ defined by
equation \ref{eqn:ipr}. However a simple trick produces a unique
(necessarily weak) solution in both cases, provided an additional
constraint is placed on $\kappa$:

\indent Assumption C: There exists $\epsilon >0$ so that $\kappa =
\kappa_0 \in \bR^+$ in $\{\bx: |z-z_s|<\epsilon\}$.

\noindent That is, $\kappa$ is constant near the plane $z=z_s$. See
Appendix \appSingular for a proof of this statement.

Define the coefficient-to-data map (``forward map'', ``modeling
operator'') $S$ by
\begin{equation}
\label{eqn:fwd}
S[\kappa]f = p|_{z=z_r}
\end{equation}
The notation reminds the reader that $S$ is linear in $f$ but
nonlinear in $\kappa$.

Theorem 2 in \cite[]{BlazekStolkSymes:13} assures that $S[\kappa]$ is
well-defined as a map from right-hand sides $f$ having
square-integrable time derivatives to square-integrable traces. With
more regularity in time, the map is also differentiable in its
dependence on $\kappa$. In Appendix \ref{appSingular}, this result is
extended to the singular sources \ref{eqn:ipr}, \ref{eqn:ss}.

In fact, stronger results hold under Assumption C. The special cases
considered below have stronger continuity properties.

After Radon transform, this source takes the form
$\tw(\xi,t)\delta(z)$, where $\xi$ is the plane wave slowness.

@@

\section{Plane Wave Modeling}

@@

\append{Singular RIght-hand Sides and Weak Solutions}
Theorem 1 in \cite[]{BlazekStolkSymes:13} provides a strong solution of the system \ref{eqn:awe}: $p$, respectively $\bv$, is continuous
in time with values in the Sobolev space $H^1(\bR^d)$, respectively
$H^1_{\rm div}(\bR^d)$ and
also continuously differentiable with values in $L^2(\bR^d)$,
respectively $(L^2(\bR^d))^d$. Therefore all of the terms in
\ref{eqn:awe} are square-integrable functions.

Theorem 4 in \cite[]{BlazekStolkSymes:13} implies that for each $t$, $p(\cdot,t)$ and $\bv(\cdot,t)$
have compact support, and that the support expands in $t$ at the rate
bounded by the sup of the velocity $c=\sqrt{\kappa/\rho}$, as one
would expect. Therefore, for any fixed max time $T$, it is possible to
treat the problem for $0\le t \le T$ as a Dirichlet problem in a
domain $\Omega$ of sufficiently large diameter. Without loss of
generality, assume that $\mbox{supp }f \subset \Omega \times [0,T]$.

The isotropic point radiator \ref{eqn:ipr} is not
square-integrable, so the theory in \cite[]{BlazekStolkSymes:13} does
not apply directly to the system
\begin{eqnarray}
\label{eqn:aweptsrc}
\frac{\partial p}{\partial t} + \kappa \nabla \cdot {\bf v} &=&
                                                                w(t)\delta(\bx-\bx_s) \nonumber\\
\rho \frac{\partial {\bf v}}{\partial t} + \nabla p = 0 \nonumber\\
p,{\bf v}&=&0, t \ll 0
\end{eqnarray}
However there is a simple trick that makes this
theory applicable, provided one assumes that $\kappa$ is constant
($=\kappa_0 \in \bR^+$) near
$\bx=\bx_s$ (Assumption C). With this constraint, the problem
\ref{eqn:aweptsrc} may be shown to have a unique causal weak solution.,

Note that the existence of weak solutions for problems of this type
has been 
well-known for decades, with the additional assumption that the coefficients are
smooth, see for example \cite{Lax:PDENotes}. Using the progressing
wave expansion of the fundamental solution for smooth coefficient
problems (for example, \cite{Friedlander:75}) in place of the analytic
expression for the constant-coefficient point radiator solution given
below, it is possible to obtain a similar result under the weaker
\begin{itemize}
\item[Assumption C']: $\kappa$ is smooth near $\bx_s$.
\end{itemize}

Denote by $(p_0,\bv_0)$ the (weak) solution of the
system \ref{eqn:awe} with $\kappa$ replaced by $\kappa_0$. For $d=3$,
\[
 p_0(\bx,t) = \frac{1}{4\pi |\bx-\bx_s|}w\left(t-\frac{ |\bx-\bx_s|}{c_0}\right)
\]
in which $c_0=\sqrt{\kappa_0/\rho_0}$. There is a similar expression
for $\bv_0$, and for other space dimensions $d$. Because $w$ is smooth, both $p$ and $\bv$ are smooth
except at $\bx=\bx_s$. Pick $\phi \in C_0^{\infty}(\bR^d)$ so that
$\phi = 1$ in a neighborhood of $\bx_s$ and $\kappa=\kappa_0$ on
$\mbox{supp }\phi$. Then
\begin{eqnarray}
\label{eqn:awecut}
\frac{\partial (\phi p_0)}{\partial t} + \kappa \nabla \cdot (\phi {\bf v}_0) &=&
                           w(t)\delta(\bx-\bx_s) + f_0  \nonumber\\
\rho \frac{\partial (\phi {\bf v}_0)}{\partial t} + \nabla (\phi p_0) = {\bf g}_0 \nonumber\\
\end{eqnarray}
in which $f_0$ and ${\bf g}_0$ are smooth of compact support. Denote by
$(p_1,\bv_1)$ the strong solution of
\begin{eqnarray}
\label{eqn:awesupp}
\frac{\partial p_1}{\partial t} + \kappa \nabla \cdot {\bf v}_1 &=&
                                                                -f_0 \nonumber\\
\rho \frac{\partial {\bf v}_1}{\partial t} + \nabla p_1 = -{\bf g}_0 \nonumber\\
p_1,{\bf v}_1&=&0, t \ll 0
\end{eqnarray}
as provided by Theorem 1 of \cite{BlazekStolkSymes:13}. Then
\begin{eqnarray*}
  p & = & \phi p_0 + p_1\\
  \bv & = & \phi \bv_0 + \bv_1
\end{eqnarray*}
defines the unique weak solution of the system \ref{eqn:aweptsrc}.

\cite{SantosaSymes:00} use an analogous construction for the
second-order wave equation to give $L^2$ bounds on the point radiator
solution.

As mentioned in the Preliminaries section, a similar argument
establishes the existence of weak solutions for the surface source problem
\begin{eqnarray}
\label{eqn:awess}
\frac{\partial p}{\partial t} + \kappa \nabla \cdot {\bf v} &=&
                                                                w(x,y,t)\delta(z-z_s) \nonumber\\
\rho \frac{\partial {\bf v}}{\partial t} + \nabla p = 0 \nonumber\\
p,{\bf v}&=&0, t \ll 0
\end{eqnarray}
under Assumption D, that $\kappa=\kappa_0$ near $z=z_s$. An analytic
solution for $\kappa \equiv \kappa_0$ follows from that for the point
radiator problem, by superposition. Then a similar cutoff construction
and appeal to Theorem 1 of \cite{BlazekStolkSymes:13} proveds a weak
solution to \ref{eqn:awess}.

This construction also works via recourse to the progressing wave
expansion for the weaker
\begin{itemize}
\item[Assumption D']: $\epsilon > 0$ exists for which $\kappa$ is
smooth in the collar neigborhood $\{\bx: |z-z_s|<\epsilon\}$.
\end{itemize}

\append{1D Radiation Problem}
Begin with the 1D acoustics point source system. 
\begin{eqnarray}
\label{eqn:awe1d}
\frac{\partial p}{\partial t} +\kappa\frac{\partial 
  v}{\partial z} &=& w(t)\delta(z-z_s) \nonumber\\
\rho \frac{\partial v}{\partial t} + \frac{\partial p}{\partial 
  z}&=&0\nonumber\\
 p,v&=&0, t \ll 0. 
\end{eqnarray}
Since the right hand side is singular, so is the solution, so it must
be a solution in the weak sense; since it is causal, it is uniquely
determined by the system above. It follows from the weak solution
conditions that the pressure is continuous at $p=0$, whence $v$ must
have a discontinuity.

Now assume that $\kappa, \rho$ are constant. 
In $z \ne z_s$, the right hand side 
vanishes, so the solution must be locally a combination of plane
waves; causality implies that
\[
p(z,t)=a\left(t \mp \frac{z-z_s}{c}\right), \, v(z,t)=\pm b\left(t \mp
  \frac{z_s}{c}\right)
\]
From the second dynamical equation (Newton's law) it follows that $b =
a/(\rho c)$. The singularity on the LHS of the first dynamical
equation (constitutive law) is
\[
\kappa [v]_{z=z_s}\delta(z-z_s) =
2\frac{\kappa}{c}b\delta(z-z_s) = 2a\delta(z-z_s).
\] 
This must in turn equal the RHS of the constitutive law, whence
$a=w/2$. Thus
\begin{eqnarray}
\label{eqn:sol1d}
p(z,t) &=& \frac{1}{2}w\left(t \mp \frac{z-z_s}{c}\right) \nonumber \\
v(z,t) &=& \pm\frac{1}{2\rho c}w\left(t \mp \frac{z-z_s}{c}\right)
           \nonumber \\
\pm z-z_s &>& 0.
\end{eqnarray}

@@

If $\kappa, \rho $ are constant, it is possible to develop explicit
expressions for the solution of this system. Only $\kappa$ will be
subject to perturbation in any of the problems considered in this
report, so assume that that $\delta\rho=0$.
